```{r setup4, echo=FALSE}
library(stringr)
library(infer)
library(nycflights13)
library(broom)
library(dplyr)
library(ggplot2)
library(stringr)
library(rvest)
library(mosaicData) 
library(ggmosaic)
library(NHANES)
library(mosaic)
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
```

# Random Variability

Up until this point, we have thought about 

1. Data collection process (sampling and study design) and Data Quality (issues of bias)
2. Data visualization (the first step of any data analysis)
3. Modeling (to explain observed variation)

Throughout the past three chapters, we have also sprinkled in the idea that **the sample we observe is one random representation of the true population or phenomenon.** Therefore, any numerical summary of the sample, any **statistic**, would be an estimate the true numerical summary of the population, the corresponding unknown **parameter**.


If we could repeat the random sampling process, each sample we would get would be slightly different. The individual composition would differ every time. Sometimes we would randomly over represent one group people and another time we would randomly over represent another group of people, just by chance. Our statistic that we calculate would change slightly for every sample. 

If we could repeat the randomization process in an experiment, each treatment would be slightly different. The individual composition would differ every time. Sometimes we would randomly over represent one group people and another time we would randomly over represent another group of people, just by chance.  Our statistic that we calculate and compare between groups would change slightly for every reshuffling of individuals. 


This is the idea of **random variability** due to random sampling or randomization, typically. The sample composition varies and therefore, the statistics varies. 

Let's explore this concept a bit more before we formally talk about probability and chances. 


## Random Sampling from a Population

The data set we will work with contains ALL flights leaving New York City in 2013. This data represents a full census of the target population of flights leaving NYC in a particular year. 

We'll start by creating two new variables, `season` defined as winter (Oct - March) or summer (April - Sept) and `day_hour` defined as morning (midnight to noon) or afternoon (noon to midnight).

```{r}
data(flights)
flights <- flights %>% 
  na.omit() %>%
  mutate(season = case_when(
    month %in% c(10:12, 1:3) ~ "winter",
    month %in% c(4:9) ~ "summer"
  )) %>% 
  mutate(day_hour = case_when(
    between(hour, 1, 12) ~ "morning",
    between(hour, 13, 24) ~ "afternoon"
  )) %>% 
  select(arr_delay, dep_delay, season, day_hour, origin, carrier)
```

Since we have the full population, we could just describe the flights that happened. Instead, we are going to use this population to help us think about sampling variability. 

Let's take one random sample of 100 flights from the data set, using a simple random sample strategy.  Let's look at the arrival delay (in minutes) with a histogram and calculate the median and mean arrival delay.

```{r}
flights_samp1 <- flights %>% 
  sample_n(size = 100) 

flights_samp1 %>% 
  ggplot(aes(x = arr_delay)) +
  geom_histogram() +
  theme_minimal()

flights_samp1 %>%
  summarize(medians = median(arr_delay), means = mean(arr_delay))
```

```{block type='reflect'}
At this point, we haven't looked at the entire population of flights. Based on a sample of 100 flights, what do you think the distribution of arrival delays looks like for the entire population? Shape? Center? Spread?
```


Now, let's take another random sample of 100 flights from the full population of flights. 


```{r}
flights_samp2 <- flights %>% 
  sample_n(size = 100) 

flights_samp2 %>% 
  ggplot(aes(x = arr_delay)) +
  geom_histogram() +
  theme_minimal()

flights_samp2 %>%
  summarize(medians = median(arr_delay), means = mean(arr_delay))
```

```{block type='reflect'}
How does the second sample differ from the first sample? What do they have in common?
```

We could keep the process going--take a sample of 100 flights, look at the histogram and calculate the median and mean. Repeat many, many times. 

We can add a little bit of code to help us simulate this sampling 1000 times and calculate the median and mean for each sample of 100 flights. 


```{r}
sim_data <- mosaic::do(1000)*( 
  flights %>% 
    sample_n(size = 100) %>%  #Generate
    summarize(medians = median(arr_delay), means = mean(arr_delay))  #Calculate
)
```

Now we have 1000 medians and 1000 means, each corresponding to a sample median and sample mean of 100 random flights from the population. Let's summarize and visualize this simulation.

```{r}
#Summarize
summary_sim <- sim_data %>% 
  summarize(mean_medians = mean(medians), 
    mean_means = mean(means), 
    sd_medians = sd(medians), 
    sd_means = sd(means))

#Visualize
sim_data %>% 
  ggplot(aes(x = medians)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_sim, aes(xintercept = mean_medians), color = 'red') +
  theme_minimal()

#Visualize
sim_data %>% 
  ggplot(aes(x = means)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_sim, aes(xintercept = mean_means), color = 'red') +
  theme_minimal()
```

These histograms estimate the **sampling distribution** of sample median arrival delay and the **sampling distribution** of sample mean arrival delay, both of which describe the variability in the sample statistic across all possible random samples from the population. 

```{block type='reflect'}
Describe the shape, center, and spread of the sampling distribution for the sample mean.
```


###IRL: Bootstrapping 

In real life (IRL), we don't have a population to repeated draw from. We only have a sample that was already drawn from the larger target population. 

To get a sense of the sampling variability, we could try to mimic this process using our best stand-in for the population, our sample. We will call the sample our "fake population" for the moment.


####Generate

To generate the different random samples of the same size (100 flights) from our "fake population", we have to draw sample of 100 flights WITH REPLACEMENT, meaning that we have to put a flight back into the pool after drawing them out. 

```{block type='reflect'}
What would happen if we drew WITHOUT REPLACEMENT?
```


###Calculate

In our simulation above, we calculated the median and mean arrival delay. In theory, we could calculate any numerical summary of data, be it the mean, median, sd, 25th percentile, etc. 

```{r}
boot_data <- mosaic::do(1000)*( 
  flights_samp1 %>% #Start with sample
    sample_frac(replace = TRUE) %>%  #Generate by resampling with replacement
    summarize(medians = median(arr_delay), means = mean(arr_delay))  #Calculate
)
```


###Summarize

Let's summarize these 1000 medians and 1000 means generated from resampling (with replacement) from our sample (our "fake population").

```{r}
#Summarize
summary_boot <- boot_data %>% 
  summarize(mean_medians = mean(medians), 
    mean_means = mean(means), 
    sd_medians = sd(medians), 
    sd_means = sd(means))
summary_boot
```

Let's compare this to the summaries from the simulation from the population.

```{r}
summary_sim
```

They won't be exactly the same, but they should be of roughly similar magnitude. 

###Visualize

Let's visualize these 1000 medians and 1000 means generated from resampling (with replacement) from our sample (our "fake population").

```{r}
#Visualize
boot_data %>% 
  ggplot(aes(x = medians)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_boot, aes(xintercept = mean_medians), color = 'red') +
  theme_minimal()

#Visualize
boot_data %>% 
  ggplot(aes(x = means)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_boot, aes(xintercept = mean_means), color = 'red') +
  theme_minimal()
```

Let's compare this to the visuals from the simulation from the population.

```{r}
#Visualize
sim_data %>% 
  ggplot(aes(x = medians)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_sim, aes(xintercept = mean_medians), color = 'red') +
  theme_minimal()

#Visualize
sim_data %>% 
  ggplot(aes(x = means)) +
  geom_histogram(bins = 15) +
  geom_vline(data = summary_sim, aes(xintercept = mean_means), color = 'red') +
  theme_minimal()
```


The process of resampling from our sample is called **bootstrapping** and it becoming the one of main computational tools for estimating sampling variability in Statistics. 

```{block type='reflect'}
How well does bootstrapping do in mimicking the simulations from the population? What could we change to improve bootstrap's ability to mimic the simulations?
```

This is a really important concept in Statistics! We'll return to the ideas of sampling variability and bootstrapping throughout the rest of the course. 

```{block type='reflect'}
What do you think the mean arrival delay is in the population? If you had to give an interval of plausible values for the population mean, what range would you give? Why?
```


##Randomization into Groups

Now, we have been thinking about arrival delays in general. If you were planning a trip, you may be able to choose between two flights that leave at different times of day. Do morning flights have shorter arrival delays on average than afternoon flights? If so, book the early flight!

Let's look at the data! We use a random sample of 500 flights from the population to investigate this question.

```{r}
flights_samp500 <- flights %>% 
  sample_n(size = 500) 
```

Let's summarize and visualize the relationship between hour of the day (morning or afternoon) and the arrival delay.

```{r}
flights_samp500 %>%
  group_by(day_hour) %>%
  summarize(median = median(arr_delay), mean = mean(arr_delay))

flights_samp500 %>%
  ggplot(aes(x = day_hour, y = arr_delay)) +
  geom_boxplot() +
  theme_minimal()
```

```{block type='reflect'}
Based solely on the visual and numerical summary, are arrival delays shorter in the morning than in the afternoon?
```


We don't know the exact reason why some flights were scheduled in the morning or the afternoon and why one flight might be delayed (it is probably due to complex combination of things). 

If there were **no difference** between morning and afternoon flights, then the it wouldn't matter whether the flight was in the morning or afternoon. The `day_hour` variable would be a irrelevant to the arrival delay. If that were true, then we could reshuffle the values of `day_hour` and it wouldn't change our conclusions. 

Wouldn't it be great if we could see how the mean arrival delays might change if we shuffled the flights around from the "morning" group to "afternoon" group, randomly. 

In fact, wouldn't it be great if we could look at every permutation of flights between two groups.  

###IRL: Randomization Tests

In real life, we don't often consider every possible permutation (reshuffling of group members) due to the magnitude of permutations. However, we can randomly reshuffle flights about 1000 times to try to approximate many permutations. 

####Hypothesis

Our **null hypothesis** (a hypothesis that is convservative) is that there is no relationship between time of day and the arrival delay. 

####Generate

We can generate 1000 new data sets based on randomly reshuffling the labels of `day_hour`. For each of these data sets, we calculate the difference in mean arrival delay between morning and afternoon groups. 

```{r}
require(infer)
null_dist <-  flights_samp500 %>% 
    specify(arr_delay ~ day_hour) %>%
    hypothesize(null = 'independence') %>% #Hypothesize
    generate(reps = 1000, type = 'permute') %>% #Generate
    calculate(stat = 'diff in means', order = c('morning','afternoon'))
```

####Visualize

The histogram below shows the histogram of differences in means if the null hypothesis were true. The vertical line shows the observed difference in means. 

```{r}
obs <- flights_samp500 %>% 
    specify(arr_delay ~ day_hour) %>%
    calculate(stat = 'diff in means', order = c('morning','afternoon'))

null_dist %>% 
  ggplot(aes(x = stat)) +
  geom_histogram() + 
  geom_vline(data = obs, aes(xintercept = stat), color = 'red') +
  theme_minimal()
```


```{block type='reflect'}
Do you think that the mean arrival delay is different for morning and afternoon? Is the observed difference in means likely to have occurred if there were no relationship?
```